 

### 2021-03-08 vertical-thermal-review.Rmd

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(3) leaf temperature ($T_{leaf}$);
(2) the leaf traits most strongly influence leaf temperature and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including scaling across space and time.*

@@

### Light environments
**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. 
The light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b].

In heterogeneous canopies [with high gap fractions], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]

Tropical forests with densely packed overstory trees limit light to subsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [@waySunflecksTreesForests2012]

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

@@

*  However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al. 2008].

* Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020]. However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.

@@

@@

##### *(lifespan)*
*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

##### *(Deciduousness)*

@@

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*

@@

@@

# Implications

>>>>

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic [replace synthetic w comprehensive?] understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(3) leaf temperature ($T_{leaf}$);
(2) the leaf traits most strongly influence leaf temperature and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including scaling across space and time.*

@@

### Light environments

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. 
The light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b].

In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995). Lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]

Tropical forests with densely packed overstory trees limit light reaching subsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [@waySunflecksTreesForests2012]

@@

RH, in combination with temperature, determines vapor pressure deficit ($VPD$), an important metric expressing the driving force of water loss from a leaf. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from changes in light availability. However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area-based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al. 2008]. In a temperate Acer saccharum forest, ~~within-canopy Acer saccharum exhibited greater~~LMA and $N_{area}$ were greater with height early in the growing season as a result of hydrostatic constraints on leaf morphology. Later in the growing season, LMA and $N_{area}$ continued to increase in the upper canopy, but due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016]. In addition, carbon isotope composition has also been shown to increase with height, independent of light [@cobleLightAcclimationOptimizes2015]

Height-based leaf trait variation in two tall temperate species, *Quercus alba* and *Quercus velutina*, showed decreases in lobation with increased height, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020]. However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus species.

@@

Several studies point towards differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

##### *Deciduousness*

@@

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*

@@

## Implications [ & future research directions?]

*Elsa: If emphsaize future research directions, could move the remote sensing opportunities to this section and frame modeling section around opportunities to use models to further explore open questions, scale these findings in space and time, and influence predictions*
*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*

@@
